---
title: "Health and Economic Impact of Storms in United States, 1950-2011"
author: "Javier Carrasco"
date: "31/08/2017"
output: html_document
---

## Synopsis

The goal of the assignment is to explore the [NOAA Storm Database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
and investigate which types of severe weather events are most harmful on:

1. Population health (injuries and fatalities) 
2. Economy (property and crop damages)

The events in the database start in the year 1950 and end in November 2011.

## Data Processing

Load libraries.

```{r}
library(dplyr)
library(ggplot2)
```

Download and unzip the storm data file.

```{r}
dataFile = "repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists(dataFile)){
        fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(fileURL, zipFile, method="curl")
}  
```

Load storm data (only relevant columns for the analysis).

```{r}
storm <- read.csv(dataFile)[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
head(storm, 10)
```

Discard events with non casualties or property / cost damages:

```{r}
storm <- storm %>% filter(INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)
```


Add casualties column (sum of fatalities and injuries).

```{r}
storm$casualties <- storm$FATALITIES + storm$INJURIES
```

Define function to convert property / crop damage exponents to numeric values for cost calculations.

```{r}
exponentMap <- function(exponent) {
        if (exponent == "")
                return (10^0)
        map <- c("0" = 10^0, "1" = 10^1, "2" = 10^2, "3" = 10^3,
          "4" = 10^4, "5" = 10^5, "6" = 10^6, "7" = 10^7,
          "8" = 10^8, "9" = 10^9, "h" = 10^2, "k" = 10^3,
          "m" = 10^6, "b" = 10^9, "-" = 10^0, "?" = 10^0,
          "+" = 10^0)
        return (map[tolower(exponent)])
} 
```

Add economic cost column (sum of property and crop costs).

```{r}
storm$economicCost <- storm$PROPDMG * unlist(lapply(storm$PROPDMGEXP, exponentMap)) + storm$CROPDMG * unlist(lapply(storm$CROPDMGEXP, exponentMap))
```

Group casualties by event type.

```{r}
casualtiesByEvtype <- storm %>%
        group_by(EVTYPE) %>%
        summarise(casualties = sum(casualties))
head(casualtiesByEvtype, 10)
```

Group economic cost by event type.

```{r}
economicCostByEvtype <- storm %>%
        group_by(EVTYPE) %>%
        summarise(economicCost = sum(economicCost))
head(economicCostByEvtype, 10)
```

## Results

Top 20 types of storm events which are most harmful to population health.

```{r}
n <- 20
topCasualtiesByEvtype <- slice(casualtiesByEvtype, order(-casualtiesByEvtype$casualties)[1:n])

ggplot(topCasualtiesByEvtype, aes(x=reorder(EVTYPE, -casualties), y = casualties)) +
        geom_bar(stat="identity", fill = "indianred") +
        geom_text(aes(label=paste(sprintf("%d", casualties))), position=position_dodge(width=0.9), vjust=-0.25, size=3) +
        labs(title = "Top 20 most harmful storm event types to population health", x = "Event Type", y = "Casualties") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme(plot.title = element_text(hjust = 0.5))
```

Top 20 types of storm events with greatest economic impact.

```{r}
n <- 20
topEconomicCostByEvtype <- slice(economicCostByEvtype, order(-economicCostByEvtype$economicCost)[1:n])

ggplot(topEconomicCostByEvtype, aes(x=reorder(EVTYPE, -economicCost), y = economicCost/10^9)) +
        geom_bar(stat="identity", fill = "indianred") +
        geom_text(aes(label=paste(sprintf("%.02f", economicCost/10^9))), position=position_dodge(width=0.9), vjust=-0.25, size=3) +
        labs(title = "Top 20 most harmful storm event types to economy", x = "Event Type", y = "Cost in US$ billions") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme(plot.title = element_text(hjust = 0.5))
```
